---
title: "Swedish Pharmacies 2024"
output: 
  flexdashboard::flex_dashboard:
    theme:
      version: 4
      bg: "white"
      fg: "#003540"
      navbar-bg: "black"
      primary: "#005966"
      base_font: 
        google: Inter
      heading_font:
        google: Overpass
      code_font:
        google: 
          family: JetBrains Mono
          local: false
    vertical_layout: fill
editor_options: 
  chunk_output_type: console
runtime: shiny
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(leaflet)
library(shiny)
library(sf)
library(shinyWidgets)


# Load data --------------------------------------------------------------------
df_apotek <- readRDS("data/processed/df_apotek.rds")

# Load filter choices
filter_choices <- readRDS("data/processed/filter_choices.rds")
lan <- filter_choices$lan
kommungrupp <- filter_choices$kommungrupp
aktorer <- filter_choices$aktorer

# Load shape files for map polygons
lan_map <- st_read("data/processed/lan_map", quiet = TRUE)
kommun_map <- st_read("data/processed/kommun_map", quiet = TRUE)

# Colour palette for pharmacy operators (to be customised)
pharmacy_colours <- c("#ff0000", "#44ad2c", "#102d69", "lightblue", "#ff9e17")
```

Map
==================

Sidebar {.sidebar data-width=270 .my-class}
---------------------
```{r sidebar, results='asis'}
pickerInput(inputId = "select_indelning", label =  HTML("<strong>Select county:</strong>"),
            choices = list(All = "Riket",
                           County = lan),
            options = list(`style` = "background: white"))

pickerInput("select_kommungrupp", HTML("<strong>Select municipality group:</strong>"),
            choices = kommungrupp,
            selected = "All",
            options = pickerOptions(dropdownAlignRight = TRUE,
                                    `style` = "background: white;"))

pickerInput(inputId = "select_aktor", label = HTML("<strong>Select pharmacy operator:</strong>"),
            choices = aktorer, multiple = TRUE, selected = aktorer,
            options = pickerOptions(`style` = "background: white", `actions-box` = TRUE,
                                    selectAllText = "Select all", deselectAllText = "Deselect all"),
            choicesOpt = list(style = paste0("background: ", pharmacy_colours, "; color: white;")))

prettySwitch(inputId = "select_granser", label = HTML("<strong>National/county boundaries</strong>"),
             value = FALSE, fill = TRUE)

prettySwitch(inputId = "select_befolkningskarta", label = HTML("<strong>Municipality population</strong>"),
             value = FALSE, fill = TRUE)

numericRangeInput("select_dist_apotek", label = HTML("<strong>Distance to nearest pharmacy (km):</strong>"),
            min = 0, max = ceiling(max(df_apotek$dist_km_narmaste_gln_korvag[is.finite(df_apotek$dist_km_narmaste_gln_korvag)], na.rm = TRUE)),
            separator = "to",
            value = c(0, ceiling(max(df_apotek$dist_km_narmaste_gln_korvag[is.finite(df_apotek$dist_km_narmaste_gln_korvag)], na.rm = TRUE))))

sliderInput("select_sarbarhet2", label = HTML("<strong>Vulnerability index:</strong>"), value = c(0, 100),
            min = 0, max = 100, sep = " ", ticks = FALSE)

# custom button because the button doesn't work in Flexdashboard for some reason
downloadButtonRmd <- function (outputId, label = "Download", class = NULL, ...)  {
     tags$a(id = outputId, class = paste("btn btn-default shiny-download-link", 
        class), href = "", target = "_blank", download = NA, 
        icon("download"), label, ...)
}

downloadButtonRmd(outputId  = "downloadKarta", label = HTML("<strong>Download data</strong>"))

cat("<br>")
```

```{r filtrera_data}
# select pharmacies based on input from widgets
valda_apotek <- reactive({
  
  filtered_data <- df_apotek %>% 
    filter(
      aktor %in% input$select_aktor,
      percentil_sarbarhet >= input$select_sarbarhet2[1] & percentil_sarbarhet <= input$select_sarbarhet2[2],
      dist_km_narmaste_gln_korvag >= input$select_dist_apotek[1] & dist_km_narmaste_gln_korvag <= input$select_dist_apotek[2]
    )
  
  if (input$select_indelning %in% lan) {
    filtered_data <- filtered_data %>% filter(lan == input$select_indelning)
  }
  
  if (input$select_kommungrupp != "All") {
    filtered_data <- filtered_data %>% filter(kommungrupp == input$select_kommungrupp)
  }

  return(filtered_data)
})

# county and national boundaries based on region selection
lanskarta <- reactive({
  if(input$select_indelning == "Riket"){
    map <- lan_map %>% 
      summarise(geometry = st_union(geometry))
  }
  
  if(input$select_indelning %in% lan){
    map <- lan_map %>%
      filter(lan == input$select_indelning)
  }
  
  return(map)
})

# population map based on municipality population
befolkningskarta <- reactive({
  if(input$select_indelning == "Riket"){
    map <- kommun_map
  }

  if(input$select_indelning %in% lan){
    map <- kommun_map %>%
      filter(lan == input$select_indelning)
  }

  return(map)
})


# download selected pharmacies -------------------------------------------------
output$downloadKarta <- downloadHandler(
  filename = function() {
    str_glue("apotek_karta_{Sys.Date()}.xlsx")
  },
  content = function(file) {
    writexl::write_xlsx(valda_apotek() %>% 
                          select(-c(starts_with("kvartil_"))), file)
  }
)
```

Kolumn 
---------------------
```{r interaktiv_karta, out.width = '100%', out.height = '100%'}
# create palettes for pharmacies, agents, population ---------------------------
pal_apotek <- colorFactor(palette = pharmacy_colours, levels = aktorer)
bins <- c(0, 10e3, 20e3, 50e3, 100e3, 200e3, Inf)
pal_pop <- colorBin("YlOrRd", domain = kommun_map$pop, bins = bins)


# create leaflet map -----------------------------------------------------------
output$map <- renderLeaflet({
  
  # base map
    leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron) %>% 
    addScaleBar(position = "bottomleft",
                options = scaleBarOptions(imperial = FALSE)) |> 
    setView(lat = 62.17432013788275, lng = 14.946378696886965, zoom = 5)

})

# Update map reactively without changing zoom or view
observeEvent(eventExpr = {
    input$select_granser
    input$select_indelning
    input$select_kommungrupp
  }, handlerExpr = {
  
  leafletProxy("map") |> 
    clearGroup("granser")
  
  # Update only markers without changing view or zoom
  if(input$select_granser == TRUE) {
  leafletProxy("map") %>%
    addPolygons(data = lanskarta(),
                fill = FALSE,
                color = "black",
                weight = 2,
                group = "granser")
  }
})

observeEvent(eventExpr = {
    input$select_befolkningskarta
    input$select_indelning
  }, handlerExpr = {

  leafletProxy("map") |>
    clearGroup("befolkningskarta")

  if(input$select_befolkningskarta == TRUE) {
   leafletProxy("map") %>%
      addPolygons(data = befolkningskarta(),
                fillColor = ~pal_pop(pop),
                popup = ~paste0(str_glue("<strong>{kommun}</strong>"), "<br>",
                               str_glue("{format(pop, big.mark = ',')} inhabitants")),
                color = "black", fillOpacity = 0.5, weight = 0.5,
                group = "befolkningskarta")
  }
})

observeEvent(eventExpr = {
    input$select_aktor
    input$select_indelning
    input$select_kommungrupp
    input$select_sarbarhet2
    input$select_dist_apotek
  }, handlerExpr = {

  leafletProxy("map") |>
    clearGroup("apotek") %>%
    removeControl("legendApotek")

   leafletProxy("map") %>%
    addCircles(data = valda_apotek(),
               radius = 10,
               color = ~pal_apotek(aktor),
               popup = ~paste0(str_glue("<strong>{namn}</strong>"), "<br>",
                               str_glue("- <strong>Distance to nearest pharmacy:</strong> {round(dist_km_narmaste_gln_korvag, 1)} km"), "<br>",
                               str_glue("- <strong>Vulnerability index:</strong> {percentil_sarbarhet}"), "<br>",
                               str_glue("- <strong>Extra person-kilometres:</strong> {format(round(personkilometer_skillnad), big.mark = ',')}")),
               opacity = 1,
               fill = "black",
               fillOpacity = 1,
               weight = 8,
               group = "apotek") |> 
     addLegend(data = valda_apotek(),
              pal = pal_apotek,
              values = ~aktor,
              title = "Pharmacy operator",
              opacity = 1,
              position = "bottomright",
              layerId = "legendApotek"
              )
})

leafletOutput("map", height=900)
```

About
==================

Column
---------------------

### Overview

Interactive map visualizing Swedish pharmacy locations and accessibility metrics.

**Key Features:**

- Pharmacy locations across Sweden
- Vulnerability index measuring closure impact
- Population density by municipality
- Filtering by county, operator, and distance

**Methodology:**

Vulnerability index = Population served × Increased travel distance if pharmacy closes

Distances calculated using Vincenty ellipsoid (great-circle distance).

### Data Sources

**Pharmacy Locations**

- [Pipos](https://pipos.se/vara-tjanster/serviceanalys) (November 12, 2025)

**Population Data**

- [Statistics Sweden (SCB)](https://www.scb.se/vara-tjanster/oppna-data/oppna-geodata/statistik-pa-rutor/) - 1km grid squares (2024)

**Municipality Classifications**

- Statistics Sweden (SCB) (2023)

**Geographic Boundaries**

- swemaps2 R package

Column
---------------------

### Technical Details

**Calculations:**

- Distance method: Vincenty ellipsoid formula
- Map projection: WGS84 (EPSG:4326)
- Analysis tools: R (sf, geosphere, leaflet)

**Data Processing:**

- Nearest neighbor analysis for each pharmacy
- Population-weighted distance calculations
- Vulnerability metrics based on closure impact

### About

Portfolio project demonstrating geospatial analysis and interactive visualization.

**Technologies:**

- R Shiny & Flexdashboard
- Leaflet.js mapping
- Geospatial analysis (sf, geosphere)

*Last updated: November 2025*
